---
title: "R Project - Replicating ADH Regressions"
author:
  - Ulrich Bergmann
  - Lachlan Deer
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

In this exercise we are going to replicate a regression from Autor, Dorn, Hansen (AER, 2003): "The China Syndrome: Local Labor Market Effects of Import Competition in the United States". We want to regress the change in wage/salary (`relchg_avg_hhincwage_pc_pw`) on the percent change in import per worker (`d_tradeusch_pw`). You find the final regression coefficient in Column 2 of Panel A of Table 9 of the paper.

Let's first load the necessary packages to read data and do fancy regressions:

```{r}
library("readr")
library("tibble")
library("dplyr")
library("sandwich")
library("lmtest")
library("lfe")
```

And let's load the data like we always do:

```{r}
df = read_csv("data/adh_data.csv")
```

## 1. OLS regression

The core of the paper is looking at what happened to laborer's when theres an increase in us imports from china. 
Let's try and replicate part of Table 9 - namely the estimate from panel A column 2.

Their y variable is `relchg_avg_hhincwage_pc_pw`. 
The important x variable is decadal trade between the us and china `d_tradeusch_pw`.

1. Run that simple regression

```{r}
lm_1 = lm(relchg_avg_hhincwage_pc_pw ~ d_tradeusch_pw, data = df)
summary(lm_1)
```


2. Now add heteroskedasticity robust standard  (HC1). Hint: Use the `sandwich` and `lmtest` packages

```{r}
coeftest(lm_1, vcov = vcovHC(lm_1, type="HC1"))
```



Now we will start to add extra x variables.

3. Start by adding `t2` - a dummy variable for whether observation is in the second decade. 
Fit again with HC1 robust standard errors.

```{r}
lm_2 = lm(relchg_avg_hhincwage_pc_pw ~ d_tradeusch_pw + t2, data = df)
coeftest(lm_2, vcov = vcovHC(lm_2, type="HC1"))
```


## 2. Clustering

Let us now use clustertered standard errors instead. ADH cluster by `statefip`.
Hint: use the `felm` command from the `lfe` package

1. Run the basic regression with clustering

```{r}
felm_1 = felm(relchg_avg_hhincwage_pc_pw ~ d_tradeusch_pw + t2 | 0 | 0 | statefip, data = df)
summary(felm_1)
```


2. Add the following controls to your last regression:
    - `l_shind_manuf_cbp`
    - `l_sh_popedu_c`
    - `l_sh_popfborn`
    - `l_sh_empl_f`
    - `l_sh_routine33`
    - `l_task_outsource`
    
```{r}
felm_2 = felm(relchg_avg_hhincwage_pc_pw ~ d_tradeusch_pw + t2 +
                l_shind_manuf_cbp + l_sh_popedu_c + l_sh_popfborn +
                l_sh_empl_f + l_sh_routine33 + l_task_outsource
                | 0 | 0 | statefip, data = df)
summary(felm_2)
```
    


3. Add region fixed effects to your regression.
    - First find all variables in the dataset that start with `reg_`
    - Add these to your last regression
  
```{r}
names(select(df, starts_with("reg_")))

felm_3 = felm(relchg_avg_hhincwage_pc_pw ~ d_tradeusch_pw + t2 +
                l_shind_manuf_cbp + l_sh_popedu_c + l_sh_popfborn +
                l_sh_empl_f + l_sh_routine33 + l_task_outsource +
                reg_midatl + reg_encen + reg_wncen + reg_satl +
                reg_escen + reg_wscen + reg_mount + reg_pacif
              | 0 | 0 | statefip, data = df)
summary(felm_3)
```

  
## 3. Instrument Variables

1. Instrument `d_tradeusch_pw` with `d_tradeotch_pw_lag` in your last regression

```{r}
felm_4 = felm(relchg_avg_hhincwage_pc_pw ~ 1 + t2 +
                l_shind_manuf_cbp + l_sh_popedu_c + l_sh_popfborn +
                l_sh_empl_f + l_sh_routine33 + l_task_outsource +
                reg_midatl + reg_encen + reg_wncen + reg_satl +
                reg_escen + reg_wscen + reg_mount + reg_pacif
              | 0 | (d_tradeusch_pw ~ d_tradeotch_pw_lag) | statefip,
              data = df)
summary(felm_4)
```


2. Weight your regression by `timepwt48`

The `felm` function is a bit picky on the order of the weights. Let us first try to define weights at the end after the `data` argument like so:

```{r, error = TRUE}
felm_5 = felm(relchg_avg_hhincwage_pc_pw ~ 1 + t2 +
                l_shind_manuf_cbp + l_sh_popedu_c + l_sh_popfborn +
                l_sh_empl_f + l_sh_routine33 + l_task_outsource +
                reg_midatl + reg_encen + reg_wncen + reg_satl +
                reg_escen + reg_wscen + reg_mount + reg_pacif
              | 0 | (d_tradeusch_pw ~ d_tradeotch_pw_lag) | statefip, 
              data = df,
              weights = timepwt48)
summary(felm_5)
```

Felm didn't find `timepwt48` because it only assumes that columns are in `df` before you define `data = df`. We can solve this in two ways. 

1. A good rule is to have `data = df` as the last argument.

```{r, error = TRUE}
felm_5 = felm(relchg_avg_hhincwage_pc_pw ~ 1 + t2 +
                l_shind_manuf_cbp + l_sh_popedu_c + l_sh_popfborn +
                l_sh_empl_f + l_sh_routine33 + l_task_outsource +
                reg_midatl + reg_encen + reg_wncen + reg_satl +
                reg_escen + reg_wscen + reg_mount + reg_pacif
              | 0 | (d_tradeusch_pw ~ d_tradeotch_pw_lag) | statefip, 
              weights = timepwt48,
              data = df)
summary(felm_5)
```

2. Alternatively, you can define weights after `data = df`, but then you have to define the weights as `df$timepwft48` like so:

```{r}
felm_5 = felm(relchg_avg_hhincwage_pc_pw ~ 1 + t2 +
                l_shind_manuf_cbp + l_sh_popedu_c + l_sh_popfborn +
                l_sh_empl_f + l_sh_routine33 + l_task_outsource +
                reg_midatl + reg_encen + reg_wncen + reg_satl +
                reg_escen + reg_wscen + reg_mount + reg_pacif
              | 0 | (d_tradeusch_pw ~ d_tradeotch_pw_lag) | statefip, 
              data = df,
              weights = df$timepwt48)
summary(felm_5)
```

And now we have the numbers reported in Column 2 of Panel A of Table 9 of the paper.


