---
title: "Data Wrangling with `dplyr` - Exercises"
author: 
- Lachlan Deer
- Uli Bergmann
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The goal of this document is to give you some experience using functions from `dplyr` with a little less guidance. 
Proceed at your own pace, and we will provide solutions later on.

We want to build up some experience with is `dplyr`.

Load it:

```{r, warning=FALSE}
library(dplyr)
```

We are going to use a classic dataset from Graduate Econometrics texts that has gasoline consumption and some other variables by country for the 60's and 70's.
The data is inside the package `plm` and called `Gasoline`. 
Load it as follows:

```{r}
data(Gasoline, package = "plm")
```

Use the help to get a sense of what variable names mean in the data.

Now work through the following exercises:

1. Create a dataset `gasoline` which is a tibble
2. Create a subset of that only has data from the 1960s. Do this two ways, once where you don't use piping, `%>%`, and once where you do.
3. Create a subset that contains data from the years ranging from 1969 to 1973.
4. Create a subset that contains data for the years 1969, 1973 and 1977.
5. Create a dataset that contains only the columns country, year, lrpmg.
6. Create a dataset that does not contain the columns country, year, lrpmg.
7. Rename the column year to be called date.
8. Select all columns that start with "l".
9. Select the columns country, year, and all columns that contain the letters "car".
10. What does the function `pull()` do? Try it on the column `lrpmg`.
11. Create a grouped data set that groups the data by country.
12. Ungroup the dataset from 10.
13. Find the mean of lgaspcar by country. Call that variable `avg_lgaspcar`.
14. Return a dataset that computes the mean of lgaspcar for france.
15. Compute the mean, standard deviation, min and max of lgaspcar by country.
16. Which country has the highest average gasoline consumption.
17. Return a dataset that returns the countries with the highest and lowest average consumption.
18. Add a variable `count` to the dataset that has the number of times each country appears in the data - Is it balanced?
19. Create a meaningless dataset called `spam` that is the exponential of the sum of lgaspcar and lincomep. Also check out what happens if you replace `mutate()` with `transmute()`.
20. Create the lead and lag of lgaspcar for each row of data. Call the new columns `lead_lgaspcar` and `lag_lgaspcar`.
21. The following countries belong the to EU:
  ```{r, eval = FALSE}
  eu_countries <- c("austria", "belgium", "bulgaria", "croatia", "republic of cyprus",
                    "czech republic", "denmark", "estonia", "finland", "france", "germany",
                    "greece", "hungary", "ireland", "italy", "latvia", "lithuania", "luxembourg",
                    "malta", "netherla", "poland", "portugal", "romania", "slovakia", "slovenia",
                    "spain", "sweden", "u.k.")
  ```
  Create a variable `in_eu` in the gasoline data which takes the value `TRUE` if a country is in the EU. (Note that the case of the string will matter!)

22. Here's a different way to classify countries:
  
* Mediterranean: france, italy, turkey, greece, spain
* Central Europe: germany, austria, switzerl, belgium, netherla
* Anglosphere: canada, u.s.a. , u.k., ireland
* Nordic: denmark, norway, sweden
* asia: japan
  
  Create a new variable `region` that uses these definitions. (Hint: `case_when()`) will likely be your friend here.

23. Notice that in the country names switzerl and netherla are a little funky. Use the functions `mutate` and  `recode` to replace the name with the full country name.
24. Compute the variable `quintile` that computes which quintile of lgaspcar each country is. Do this only for 1960. Repeat this to create a variable `decile` with the appropriate definiiton.
26. Create a variable `high_consumption` that takes the value `TRUE` if lgaspcar is higher than the yearly average for a given country.

