---
title: "R Project - Replicating ADH Regressions"
author:
  - Ulrich Bergmann
  - Lachlan Deer
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Let's first load the necessary packages to read data and do fancy regressions:

```{r, eval = FALSE}
library("readr")
library("tibble")
library("sandwich")
library("lmtest")
```

And let's load the data like we always do:

```{r, eval = FALSE}
df = read_csv("data/adh_data.csv")
```

## 1. OLS regression

The core of the paper is looking at what happened to laborer's when theres an increase in us imports from china. 
Let's try and replicate part of Table 9 - namely the estimate from panel A column 2.

Their y variable is `relchg_avg_hhincwage_pc_pw`. 
The important x variable is decadal trade between the us and china `d_tradeusch_pw`.

1. Run that simple regression
2. Now add heteroskedasticity robust standard  (HC1). Hint: Use the `sandwich` and `lmtest` packages

Now we will start to add extra x variables.

3. Start by adding `t2` - a dummy variable for whether observation is in the second decade. 
Fit again with HC1 robust standard errors.

## 2. Clustering

Let us now use clustertered standard errors instead. ADH cluster by `statefip`.
Hint: use the `felm` package.

1. Run the basic regression with clustering
2. Add the following controls to your last regression:
    - `l_shind_manuf_cbp`
    - `l_sh_popedu_c`
    - `l_sh_popfborn`
    - `l_sh_empl_f`
    - `l_sh_routine33`
    - `l_task_outsource`

3. Add region fixed effects to your regression.
    - First find all variables in the dataset that start with `reg_`
    - Add these to your last regression
  
## 3. Instrument Variables

1. Instrument `d_tradeusch_pw` with `d_tradeotch_pw_lag` in your last regression
2. Weight your regression by `timepwt48`





