---
title: "R Project - Aggregate Data in ADH"
author:
  - Ulrich Bergmann
  - Lachlan Deer
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We want to look for some aggregate facts about the US economy and its trade patterns to get a sense of how some macroeconomic indicators had evolved around China's WTO accession. In particular we will show that

    - Expansion of Chinese Trade. Essentially all of US trade growth since the 1990s is from the expansion of Chinese imports.
    - Fall in Real Interest Rates Around the time the Chinese trade expanded.
    - Expansion of the Trade Deficit during this time period.

## Load Necessary Packages

Install the ones you do not have yet.

```{r, eval = FALSE}
library("fredr")
library("purrr")
library("dplyr")
library("readr")
library("tidyr")
library("ggplot2")
library("magrittr")
library("lubridate")
library("PerformanceAnalytics")
```

## Get Data from Federal Reserve

We need the following variables from FRED:


```{r, eval = FALSE}
codes = c("GDP", "IMP0015", "IMPCH", "EXP0015", "GS1", "CPILFESL")
```

We also need to tell FRED our API key to authenticate ourselves. For this course, you can use the following command:

```{r, eval = FALSE}
api_key = "d89c49825894bf6e766a2e0afd8ba4f7"
fredr_set_key(api_key)
```

1. Use the `fredr_series_observations` command on a single variable to get the data for that variable
2. Now, find the option to pull only data starting at `1990-01-01`. Make sure you format this number as a date.
3. What happens if you use the previous command on `codes` instead of a single variable name?
4. Solve it by applying the correct map function which returns a dataframe by row-binding. Save your dataset as `df_raw`

## Data Transformations

For the rest of this part, take `df_raw` and save the transformed output as `df`:

5. Now split your data into columns
6. Rename your newly created columns as such:
    - `gdp` <- `GDP`,
    - `imp_ch` <- `IMPCH`,
    - `imp_all` <- `IMP0015`,
    - `exp_all` <- `EXP0015`,
    - `t_bill` <- `GS1`,
    - `cpi` <- `CPILFESL`
7. `cpi` is coded in billions of USD while exports and imports are in millions, multiply `cpi` by `1000` to have all values in millions
8. create additional variables for our date using the `lubridate` module. We want columns `year`, `quarter`, `month`, `day` that only contain this part from the `date` column. Find the correct functions in `lubridate`  to achieve this
9. Sort your data by `date`

We see that gdp is coded quaterly while the imports and exports are per month. We need the data grouped annually and quarterly respectively for the next two parts:

## Data Grouping

10. Group `df` quarterly into a dataframe called `df_quarter`. In this dataset you want
  - `start_date` as the minimum of date,
  - `gdp`, `imp_all`, `imp_ch`, `exp_all` all aggregated as sums (how do you deal with NA's?)
  - `cpi`, `t_bill` aggregated as averages
11. Group `df` annually into a dataframe called `df_year. In this dataset you want
  - `gdp`, `imp_all`, `imp_ch` aggregated as sums (how do you deal with NA's?)

